The Form of Uncertainty Affects Selection for Social Learning

### 2023-02-26 ms.tex
final

\documentclass[letterpaper,11.5pt]{scrartcl}
\usepackage{totcount}
% \documentclass[11pt]{report}
% \documentclass{report}
% \documentclass{book}
\usepackage[bookmarks, hidelinks]{hyperref}
\usepackage{amssymb,amsmath}
\usepackage[title]{appendix}
% \usepackage{fullpage}
\usepackage{tabulary}
\usepackage{tabularx}
\usepackage{float}
% \usepackage[margin=0.50in]{geometry}
\usepackage[margin=1.00in]{geometry}
% \usepackage[modulo]{lineno}
\usepackage{lineno}
\linenumbers

@@

To analyze the effect of the four principal uncertainty factors, we systematically
varied uncertainty values %(Table~\ref{tab:modelParameters}) 
and observed how frequently social learning evolved across 1000 trial simulations for each combination of parameters tested. %We varied $u \in \{0.0, 0.1, \ldots, 1.0\}$; $\pilow \in \{0.1, 0.45, 0.8\}$; $B \in \{2, 4, 10\}$; and $L \in \{1,2,4,8\}$ for $B=2$ and $L \in \{1,B/2,B,2B\}$ for $B=4,10$.  
\edit{We observed three outcome measures: first, the social learning (SL) fixation frequency, denoted
$\meansl$, is how frequently
social learning evolved to fixation across all trials for a given parameter setting. To support our
analysis about why we observe the patterns in social learning fixation frequency that we did, we also
measured the mean number of generations to fixation, which measures the strength of selection.} Finally we
measured the average net payoffs at the end of the simulation across agents and trials, normalized by lifespan. We analyze
outcomes by plotting $\meansl$ on the y-axis and environmental variability on
x-axis, since we theoretically expect that $\meansl$ will always decrease monotonically from 1 to 0 as $u$ increases. %Note that the hypothesis-testing concept of significance is meaningless here because we could make any small outcome difference ``significant'' by running more simulation trials.  We instead study the patterns of outcome variables over systematically varied uncertainty factor values.

@@

\clearpage
\begin{figure}
  \caption{Comparison of observed average payoffs in simulations with average payoffs
    obtained by populations homogeneously initialized to be all social or all asocial learners. Often the simulated payoffs follow the payoffs from the better-performing homogeneous group, with some exceptions discussed in the main text.}
  \label{fig:payoffs}
  \vspace{1em}
  \centering 
    {\textsf A \quad Low payoff, $\pilow = 0.1$; selection-set size, $B = 2$} \\
    \begin{subfigure}[]{0.4\textwidth}
      \centering
      \includegraphics[width=\textwidth]{Figures/mean_prev_net_payoff_over_u_lowpayoff=0.1_nbehaviors=2.pdf}
      \\[-0.8em]
      {\centering \quad Environmental variability, $u$}
    \end{subfigure} \\[1.5em]
    \centering
    {\textsf B \quad Low payoff, $\pilow = 0.45$; selection-set size, $B = 4$} \\
    \begin{subfigure}[]{0.4\textwidth}
      % {\large \textsf B}
      \centering
      \includegraphics[width=\textwidth]{Figures/mean_prev_net_payoff_over_u_lowpayoff=0.45_nbehaviors=4.pdf}
      \\[-0.8em]
      {\centering \quad Environmental variability, $u$}
    \end{subfigure} \\[1.5em]
    {\textsf C \quad Low payoff, $\pilow = 0.45$; selection-set size, $B = 10$} \\
    \begin{subfigure}[]{0.4\textwidth}
      \centering
      % {\large \textsf C}
      \includegraphics[width=\textwidth]{Figures/mean_prev_net_payoff_over_u_lowpayoff=0.45_nbehaviors=10.pdf}
      \\[-0.8em]
      {\centering \quad Environmental variability, $u$}

    \end{subfigure}

@@

%We found that the 
The evolution of social learning is a complex phenomenon, dependent on many interrelated factors, yet it is widespread across a
range of taxa. Although social learning is broadly studied, few studies have taken a systematic approach to understanding the role
of uncertainty in the evolution \edit{of} social learning. Understanding the role of uncertainty in social evolution is particularly important
for understanding human evolution as evidence increasingly suggests that the genus \emph{Homo} emerged in the context of substantial
(nonstationary) environmental variability \citep{anton_etal2014, levin2015}, and our global dispersal
\edit{meant} that human groups regularly face novel environments. \edit{These conditions, combined with long lifespan, obviated the possibility of genetic adaptation to specific ecological circumstances, requiring instead the evolution of learning mechanisms for dealing with uncertainty. In our model, numerous behavioral options led agents to be more cognitively flexible via the softmax decision rule, and thus better able to adjust their behavior if social learning provided outdated information. While Rogers-style models of social learning evolution may also predict greater social learning reliance in more challenging contexts (larger selection-set sizes), these models rely on limited individual learning mechanisms that are furthermore mutually exclusive with social learning. This may prevent social learning from evolving under such high values of environmental variability as we observed here, even if these models included more adaptive individual learning mechanisms.} 
By carefully identifying and operationalizing different forms of uncertainty common in social-learning models, we developed a more thorough understanding of how uncertainty can favor social learning and how different forms of uncertainty interact to favor its evolution. By explicitly modeling a key individual-level learning mechanism shared across taxa, namely the ability to adjust behavior to uncertainty and new information, we found that social learning could evolve even if it provides outdated information. %Social learning is a critical component of problem solving among social creatures. Our work identified and systematized some important forms of uncertainty and catalogued their effects on the evolution of social learning.
Our work, then, both advances the theory of social learning and could eventually be of practical use in identifying specific mechanisms underlying adaptations to uncertain environments.

@@

\paragraph{Acknowledgements.} %\mt{Do we write any at this point? Please fill in as you see fit.} \cm{can do later since should be blinded anyway}
%

@@

\paragraph{Conflict of interest.} No conflict declared.

>>>>

\documentclass[letterpaper,11.5pt]{scrartcl}
\usepackage{totcount}
\usepackage[chatter]{rotating}
% \documentclass[11pt]{report}
% \documentclass{report}
% \documentclass{book}
\usepackage[bookmarks, hidelinks]{hyperref}
\usepackage{amssymb,amsmath}
\usepackage[title]{appendix}
% \usepackage{fullpage}
\usepackage{tabulary}
\usepackage{tabularx}
\usepackage{float}
% \usepackage[margin=0.50in]{geometry}
\usepackage[margin=1.00in]{geometry}
% \usepackage[modulo]{lineno}
\usepackage{lineno}
\linenumbers

@@

To analyze the effect of the four principal uncertainty factors, we systematically
varied uncertainty values %(Table~\ref{tab:modelParameters}) 
and observed how frequently social learning evolved across 1000 trial simulations for each combination of parameters tested. %We varied $u \in \{0.0, 0.1, \ldots, 1.0\}$; $\pilow \in \{0.1, 0.45, 0.8\}$; $B \in \{2, 4, 10\}$; and $L \in \{1,2,4,8\}$ for $B=2$ and $L \in \{1,B/2,B,2B\}$ for $B=4,10$.  
\edit{We observed three outcome measures: first, the social learning (SL) fixation frequency, denoted $\meansl$, is how frequently social learning evolved to fixation across all trials for a given parameter setting. To support our analysis about why we observed the patterns in social learning fixation frequency that we did, we also measured the mean number of generations to fixation, which measures the strength of selection.} Finally we measured the average net payoffs at the end of the simulation across agents and trials, normalized by lifespan. We analyze
outcomes by plotting $\meansl$ on the y-axis and environmental variability on
x-axis, since we theoretically expect that $\meansl$ will always decrease monotonically from 1 to 0 as $u$ increases. %Note that the hypothesis-testing concept of significance is meaningless here because we could make any small outcome difference ``significant'' by running more simulation trials.  We instead study the patterns of outcome variables over systematically varied uncertainty factor values.

@@

\clearpage
\begin{figure}
  \caption{Comparison of observed average payoffs in simulations with average payoffs
    obtained by populations homogeneously initialized to be all social or all asocial learners. Often the simulated payoffs follow the payoffs from the better-performing homogeneous group, with some exceptions discussed in the main text.}
  \label{fig:payoffs}
  \vspace{1em}
  \centering 
    {\textsf A \quad Low payoff, $\pilow = 0.1$; selection-set size, $B = 2$} \\
    % \begin{subfigure}[]{0.4\textwidth}
    \begin{minipage}[]{0.4\textwidth}
      \centering
      \begin{rotate}{90}
        {\parbox{3.0in}{
            \centering
            % \vspace{-1.0em}\hspace{-1.5em} {\huge Low payoff, $\pilow = 0.1$} \\[1em]
            {\hspace{-7em} Mean payoffs, $\meanpi / L$}
        }}
      \end{rotate}%
      \includegraphics[width=\textwidth]{Figures/mean_prev_net_payoff_over_u_lowpayoff=0.1_nbehaviors=2.pdf}
      \\[-0.8em]
      {\centering \quad Environmental variability, $u$}
    \end{minipage} \\[1.5em]
    % \end{subfigure} \\[1.5em]
    \centering
    {\textsf B \quad Low payoff, $\pilow = 0.45$; selection-set size, $B = 4$} \\
    % \begin{subfigure}[]{0.4\textwidth}
    \begin{minipage}[]{0.4\textwidth}
      % {\large \textsf B}
      \centering
      \begin{rotate}{90}
        {\parbox{3.0in}{
            \centering
            % \vspace{-1.0em}\hspace{-1.5em} {\huge Low payoff, $\pilow = 0.1$} \\[1em]
            {\hspace{-7em} Mean payoffs, $\meanpi / L$}
        }}
      \end{rotate}%
      \includegraphics[width=\textwidth]{Figures/mean_prev_net_payoff_over_u_lowpayoff=0.45_nbehaviors=4.pdf}
      \\[-0.8em]
      {\centering \quad Environmental variability, $u$}
    \end{minipage} \\[1.5em]
    % \end{subfigure} \\[1.5em]
    {\textsf C \quad Low payoff, $\pilow = 0.45$; selection-set size, $B = 10$} \\
    % \begin{subfigure}[]{0.4\textwidth}
    \begin{minipage}[]{0.4\textwidth}
      \centering
      \begin{rotate}{90}
        {\parbox{3.0in}{
            \centering
            % \vspace{-1.0em}\hspace{-1.5em} {\huge Low payoff, $\pilow = 0.1$} \\[1em]
            {\hspace{-7em} Mean payoffs, $\meanpi / L$}
        }}
      \end{rotate}%
      % {\large \textsf C}
      \includegraphics[width=\textwidth]{Figures/mean_prev_net_payoff_over_u_lowpayoff=0.45_nbehaviors=10.pdf}
      \\[-0.8em]
      {\centering \quad Environmental variability, $u$}

    \end{minipage}
    % \end{subfigure}

@@

%We found that the 
The evolution of social learning is a complex phenomenon, dependent on many interrelated factors, yet it is widespread across a
range of taxa. Although social learning is broadly studied, few studies have taken a systematic approach to understanding the role
of uncertainty in the evolution \edit{of} social learning. Understanding the role of uncertainty in social evolution is particularly important
for understanding human evolution as evidence increasingly suggests that the genus \emph{Homo} emerged in the context of substantial
(nonstationary) environmental variability \citep{anton_etal2014, levin2015}, and our global dispersal
\edit{meant} that human groups regularly faced novel environments. \edit{These conditions, combined with long lifespan, obviated the possibility of genetic adaptation to specific ecological circumstances, requiring instead the evolution of learning mechanisms for dealing with uncertainty. In our model, numerous behavioral options led agents to be more cognitively flexible via the softmax decision rule, and thus better able to adjust their behavior if social learning provided outdated information. While Rogers-style models of social learning evolution may also predict greater social learning reliance in more challenging contexts (larger selection-set sizes), these models rely on limited individual learning mechanisms that are furthermore mutually exclusive with social learning. This may prevent social learning from evolving under such high values of environmental variability as we observed here, even if these models included more adaptive individual learning mechanisms.} 
By carefully identifying and operationalizing different forms of uncertainty common in social-learning models, we developed a more thorough understanding of how uncertainty can favor social learning and how different forms of uncertainty interact to favor its evolution. By explicitly modeling a key individual-level learning mechanism shared across taxa, namely the ability to adjust behavior to uncertainty and new information, we found that social learning could evolve even if it provides outdated information. %Social learning is a critical component of problem solving among social creatures. Our work identified and systematized some important forms of uncertainty and catalogued their effects on the evolution of social learning.
Our work, then, both advances the theory of social learning and could eventually be of practical use in identifying specific mechanisms underlying adaptations to uncertain environments.

@@

\paragraph{Acknowledgements.} We thank colleagues at the Cultural Evolution
Society and Cognitive Science Society annual meetings in 2022 who provided helpful feedback
to presentations of this work. We also thank the editor and two anonymous reviewers
for helping to improve to the original manuscript.%\mt{Do we write any at this point? Please fill in as you see fit.} \cm{can do later since should be blinded anyway}
%

@@

\paragraph{Conflict of interest.} No conflict declared. \\[1em]
